The Precambrian paleogeography of Laurentia

### 2020-11-04 Manuscript/Laurentia_Paleogeo_Manuscript.tex
Update Laurentia_Paleogeo_Manuscript.tex

\subsubsection{Paleogeographic connections prior to Laurentia assembly}

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia.

@@

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia. The ca. 1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continous with the South Australia his interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny. Another challenge with interpreting data from North Australia A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.

>>>>

\subsubsection{Paleogeographic connections prior to initial Laurentia assembly}

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of lithosphere of another continent that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest that the Suns\'as orogen of Amazonia is the best match for southern and central Appalachian inliers. This positioning is consistent with a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly (Fig. \ref{fig:Grenville_reconstructions}; \citealp{Hoffman1991a, Evans2013b, Cawood2017a}). While the lack of ca. 1100 to 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; \citealp{DAgrella-Filho2021a}) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a} in conjunction with rifting in eastern Laurentia. A significantly later separation ca. 600 Ma would be predicted if Amazonia were further north (present-day coordinates) along Laurentia's margin given the lack of evidence of riding until ca. 620 Ma north of the New York promontory.

@@

Metamorphism ca. 1.6 Ga on the eastern margin of the North Australia craton associated with the Issan-Jana orogeny has been interpreted to  be result of collisional orogenesis with the western Laurentia margin \citep{Nordsvan2018a, Pourteau2018a, Gibson2020a}. That Laurentia was the conjugate continent for this orogeny is argued to be supported by detrital zircon date spectra from the Georgetown Inlier of the eastern North Australia craton that have similarities with possible Laurentia sources \citep{Nordsvan2018a}. In the model of \cite{Nordsvan2018a}, the inlier is a continental ribbon that rifted from Laurentia ca. 1.68 Ga and was then caught up in ca. 1.60 Ga collision between the North Australia craton and northwest Laurentia although others interpret it to be part of an extended Australian margin \citep{Gibson2020a}. The ca. 1.60 Ga Racklan-Forward orogeny in northwest Laurentia records arc-continent collision that could have been followed by continent-continent collision between Laurentia and Australia \citep{Thorkelson2005a, Furlanetto2013a}. This timing of the conjoinment of the cratons would put Australia in a position that honors subsequent Mesoproterozoic correlations. Detrital zircons dating from ca. 1.61 to 1.50 Ga in the ca. 1.45 Ga lower Belt-Purcell supergroup of Laurentia are interpreted to have been sourced from the North Australia craton \citep{Jones2015a}. Additionally, ca. 1.44 Ga granites in East Antarctica (recovered as glacial clasts and inferred from detrital zircons) have the same age and isotopic signatures as the distinctive `A-type' granites in Laurentia \citep{Goodge2008a}. The interpretation of \cite{Goodge2008a, Goodge2017a} is that there is an extension of the southwest Laurentian A-type magmatic belt into Antarctica (that is currently overlain by the East Antarctic ice sheet). The correlation of the eastern North Australia craton with northwest Laurentia and that of \textbf{s}outh\textbf{w}est Laurentia with \textbf{E}ast \textbf{A}n\textbf{t}arctica leads to the SWEAT fit proposed to have initiated in the Paleoproterozoic by \cite{Moores1991a}. Given that this tight-fit configuration likely was not sustained into the Neoproterozoic, as discussed below, researchers have taken to referring to this configuration as  the ``proto-SWEAT'' reconstruction \citep{Payne2009b, Kirscher2020a}. A comparison between paleomagnetic data from the ca. 1.32 Ga Derim Derim sills of the North Australia craton and the ca. 1.31 Ga Nain anorthosite of Laurentia are consistent with this SWEAT configuration leading to the interpretation that it persisted from 1.6 Ga to at least 1.3 Ga \citep{Kirscher2020a}. If the North Australian craton was continous with the South Australia his interpretation would have the ca. 1.47 to 1.40 Ga Belt-Purcell basin be an intracontinental rift and makes it more difficult to explain the ca. 1.35 Ga East Kootenay orogeny. Another challenge with interpreting data from North Australia A paleomagnetic pole from the ca. 1.21 Ga Gnowangerup?Fraser dike swarm is argued to be inconsistent with a conjoined relationship between Australia and Laurentia as it implies a high-latitude for Australia and distinct shape of the apparent polar wander path\citep{Pisarevsky2014a}. There is not a coeval pole from Laurentia for comparison at ca. 1.21 Ga and data do indicate poleward motion for Laurentia between the ca. 1.24 and 1.18 Ga constraints. Nevertheless, comparison between latest Mesoproterozoic (ca. 1070 Ma) and Neoproterozoic (ca. 750 Ma) paleomagnetic poles from Australia and Laurentia are inconsistent with a configuration where Australia+East Antarctica are both tight against western Laurentia as they are in the proto-SWEAT fit. Rather, while these pole comparisons could be consistent with East Antarctica against southwest Laurentia, they require the eastern Australian margin to be rotated further away from Laurentia as in Figure \ref{fig:Grenville_reconstructions}. There is a similarity between Australia's paleomagnetic pole database and that of Laurentia's in that both pole paths have a similar position between ca. 1070 Ma and ca. 770 Ma poles \citep{Swanson-Hysell2012a}. This similarity could support interpretations of a unified Rodinia containing both Australia and Laurentia throughout that time interval \citep{Swanson-Hysell2012a} that subsequently broke up ca. 650 Ma \citep{Li2011a}.
